Refunctionalization of Abstract Abstract Machines (Functional Pearl)

### 2018-03-17 paper.tex
update

%% Title information
%\title{Pushdown Control Flow Analysis via Refunctionalization (Functional Pearl)}         %% [Short Title] is optional;
\title{Refunctionalization of Abstract Abstract Machines (Functional Pearl)}         %% [Short Title] is optional;
                                        %% when present, will be used in
                                        %% header instead of Full Title.
%\titlenote{with title note}             %% \titlenote is optional;
                                        %% can be repeated if necessary;
                                        %% contents suppressed with 'anonymous'
\subtitle{Filling the Gap Between Abstracting Abstract Machines and Abstracting Definitional Interpreters}        %% \subtitle is optional
%\subtitlenote{with subtitle note}       %% \subtitlenote is optional;
                                        %% can be repeated if necessary;
                                        %% contents suppressed with 'anonymous'

@@

  In this paper, we show their correspondence and how to syntactically transform small-step
  abstract abstract machines into big-step abstract definitional
  interpreters.
  The transformations include linearization, fusing, disentangling, refunctionalizing,
  and monadification (or un-CPS to direct-style). Linearization expresses
  non-deterministic choices by first-order data types, after which refunctionalization
  sequentializes the evaluation order by higher-order functions.
  All transformations properly handle the collecting semantics and the
  nondeterminism of abstract interpretation.
  %After each transformation, we also obtain an intermediate form of abstract interpreter.

@@

\textbf{Nondeterminism Abstraction.}
To handle the inevitable nondeterminism, we define a generic function @nd@ as follows:
\begin{lstlisting}
def nd[T,S](ts: Set[T], acc: S, f: (T, S, S => S) => S, g: S => S): S = {
  if (ts.isEmpty) g(acc)
  else f(ts.head, acc, (vss: S) => nd(ts.tail, vss, f, g))
}
\end{lstlisting}

@@

\textbf{Evaluation without Caching.}
Since our first version of @aeval@ implements the nondeterministic
evaluation, given the defined @nd@, the evaluation function @aeval@ can be easily
defined. In the case that @e@ is a @let@ expression,
the overall idea is to perform a two-step, depth-first evaluation. \todo{refine}

@@

\textbf{Call/Return Matching through Refunctionalization and Caching.}
\todo{how refunc and caching help}

@@

As previously identified, there are two layers of continuations: one for the normal stack of the analyzed language, and the other for the nondeterministic choices of closures.
This corresponds to the extended continuation-passing style.
But in our defining language (Scala), the delimited controls rely on the continuation-passing transformation which is only one layer.
To address this problem, based on the @nd@ operator we have before, we introduce another operator @ndcps@ for handling non-deterministic choices.

The @ndcps@ itself is implemented with delimited controls.
The function takes two arguments: a set of elements with type @T@, and an initial accumulated value of type @Set[S]@.
The @ndcps@ function iteratively returns a value of type @T@ with annotated by
\texttt{\@ cps[Set[T]]} \note{JD: Tiark, can you fix the at sign? I fixed it for the other paper, but my fix doesn't work here} that denotes the final returned value of @ndcps@ is type @Set[T]@.
The @f@ introduced by the @shift@ operator is the delimited continuation at the call-site of @ndcps@.

\begin{lstlisting}
def nd[T,S](ts: Set[T], acc: Set[S], k: T => Set[S]): Set[S] = {
  if (ts.isEmpty) acc
  else {
    val vss = k(ts.head)
    nd(ts.tail, acc ++ vss, k)
  }
}

def ndcps[T,S](ts: Set[T], acc: Set[S]): T @cps[Set[S]] = shift { f: (T => Set[S]) =>
  nd(ts, acc, f)
}
\end{lstlisting}

@@

\begin{lstlisting}
def aeval(e: Expr, env: Env, store: BStore, time: Time): Ans @cps[Ans] = {
  val new_time = (e::time).take(k)
  e match {
    case Let(x, App(f, ae), e) =>
      val closures = atomicEval(f, env, store).asInstanceOf[Set[Clos]]
      val Clos(Lam(v, body), c_env) = ndcps[Clos, VS](closures, Set[VS]())
      val vbaddr = allocBind(v, new_time)
      val new_cenv = c_env + (v -> vbaddr)
      val new_store = store.update(vbaddr, aeval(ae, env, store))
      val bodyvss = aeval(body, new_cenv, new_store, new_time)
      val VS(vals, time, store) = ndcps[VS, VS](bodyvss, Set[VS]())
      val xbaddr = allocBind(x, time)
      val new_env = env + (x -> xbaddr)
      val new_vsstore = store.update(xbaddr, vals)
      aeval(e, new_env, new_vsstore, time)

@@

The function @analyze@ is also changed by adding a @reset@ around the function call to @aeval@.

@@

This paper is greatly inspired by~\citeauthor{darais2017abstracting}'s work.
The difference on the surface is that ours works for A-Normal Form $\lambda$-calculus,
instead of plain $\lambda$-calculus, although our work can also be easily adopted to handle a plain
$\lambda$-calculus.
But except of simply showing the final form of abstract definitional interpreters,
we reveal that refunctionalization plays an important role for inheriting the
stack from the metalanguage.
The target of the series of transformations in this paper is abstract definitional
interpreters, and we additionally demonstrate abstract interpreters written direct-style with
delimited controls. The correspondence of delimited controls (as well as continuations) and monads
are well-known \cite{Danvy:1990:AC:91556.91622, wadler1992essence, danvy1992representing, moggi1991notions}.

\section{Conclusion}\label{sec:conclusion}
\note{todo}

>>>>

%% Title information
%\title{Pushdown Control Flow Analysis via Refunctionalization (Functional Pearl)}         %% [Short Title] is optional;
\title{Refunctionalization of Abstract Abstract Machines (Functional Pearl)}         %% [Short Title] is optional;
                                        %% when present, will be used in
                                        %% header instead of Full Title.
%\titlenote{with title note}             %% \titlenote is optional;
                                        %% can be repeated if necessary;
                                        %% contents suppressed with 'anonymous'
\subtitle{Filling the Gap Between Abstract Abstract Machines and Abstract Definitional Interpreters}        %% \subtitle is optional
%\subtitlenote{with subtitle note}       %% \subtitlenote is optional;
                                        %% can be repeated if necessary;
                                        %% contents suppressed with 'anonymous'

@@

  In this paper, we show their correspondence and how to syntactically transform small-step
  abstract abstract machines into big-step abstract definitional
  interpreters.
  The transformations include linearization, fusing, disentangling, refunctionalizing,
  and un-CPS to direct-style with delimited controls. Linearization expresses
  non-deterministic choices by first-order data types, after which refunctionalization
  sequentializes the evaluation order by higher-order functions.
  All transformations properly handle the collecting semantics and the
  nondeterminism of abstract interpretation.
  %After each transformation, we also obtain an intermediate form of abstract interpreter.

@@

\textbf{Nondeterminism Abstractions.}
To handle the inevitable nondeterminism, we define a generic function @nd@ as follows:
\begin{lstlisting}
def nd[T,S](ts: Set[T], acc: S, f: (T, S, S => S) => S, g: S => S): S = {
  if (ts.isEmpty) g(acc)
  else f(ts.head, acc, (vss: S) => nd(ts.tail, vss, f, g))
}
\end{lstlisting}

@@

\textbf{Evaluation without Caching.}
Since our first version of @aeval@ implements the nondeterministic
evaluation, given the defined @nd@, the evaluation function @aeval@ can be easily
defined. In the case that @e@ is a @let@ expression,
we invoke @nd@ for twice and perform a nesting two-step, depth-first evaluation
over possible values of @App(f, ae)@.

@@

\textbf{Call/Return Matching through Refunctionalization and Caching.}
The refunctionalized AAM we obtained has the perfect match on return-flows even though we do not have a stack model. The higher-order functions representing continuations already connect all the execution in order. In fact, the continuations (call stack) of the analyzed language is blended into the call stack of our defining language (Scala) through refunctionalization, then the call/return of the analyzed language is naturally matched.

In fact, what we started is an AAM with an unbounded stack which already precisely matches the calls and returns.
Refunctionalization makes the interpreter be a ``big-step'' semantics artifact, then the consequence is we have no places to save the context information. 
But refunctionalization not only forces us to sequentialize the order non-deterministic evaluation through higher-order functions representing continuations, 
but also drive us to use a new caching algorithm.

Indeed, the caching algorithm plays an important role to make the analysis computable.

An interesting implication of this is if we apply the caching algorithm with some necessary changes to small-step abstract machines with an unbounded stack, we are also able to establish a computable and precise call/return match.

@@

As previously identified, there are two layers of continuations: one for the normal stack of the analyzed language, and the other for the nondeterministic choices of closures.
This corresponds to the extended continuation-passing style.
But in our defining language (Scala), the delimited controls rely on the continuation-passing transformation which is only one layer.

To address this problem, we introduce two operators @nd@ and @ndcps@ for handling non-deterministic choices.

\textbf{Nondeterminism Abstractions, Again.}
The first operator @nd@ is simplified from the previous one. 
Now @nd@ only takes a function @f@ that applies on elements of type @T@, and does recursive call on itself over elements in set @ts@.

\begin{lstlisting}
def nd[T,S](ts: Set[T], acc: Set[S], f: T => Set[S]): Set[S] = {
  if (ts.isEmpty) acc
  else nd(ts.tail, acc ++ vss, f(ts.head))
}
\end{lstlisting}

The second operator @ndcps@ is implemented with delimited controls but simply calls our first operator @nd@.
\begin{lstlisting}
def ndcps[T,S](ts: Set[T], acc: Set[S]): T @cps[Set[S]] = shift { f: (T => Set[S]) =>
  nd(ts, acc, f)
}
\end{lstlisting}

The function @ndcps@ takes two arguments: a set of elements with type @T@, and an initial accumulated value of type @Set[S]@.
The @ndcps@ function iteratively returns a value of type @T@ with annotated by
\texttt{\@ cps[Set[T]]} \note{JD: Tiark, can you fix the at sign? I fixed it for the other paper, but my fix doesn't work here} that denotes the final returned value of @ndcps@ is type @Set[T]@.
The @f@ introduced by the @shift@ operator is the delimited continuation at the call-site of @ndcps@.

@@

\begin{lstlisting}
def aeval(e: Expr, env: Env, store: BStore, time: Time): Ans @cps[Ans] = {
  val new_time = (e::time).take(k)
  e match {
    case Let(x, App(f, ae), e) =>
      val closures = atomicEval(f, env, store).asInstanceOf[Set[Clos]]
      val Clos(Lam(v, body), c_env) = ndcps[Clos, VS](closures, Set[VS]())
      val vbaddr = allocBind(v, new_time)
      val new_cenv = c_env + (v -> vbaddr)
      val new_cstore = store.update(vbaddr, aeval(ae, env, store))
      val bodyvss = aeval(body, new_cenv, new_cstore, new_time)
      val VS(vals, time, vsstore) = ndcps[VS, VS](bodyvss, Set[VS]())
      val xbaddr = allocBind(x, time)
      val new_env = env + (x -> xbaddr)
      val new_store = vsstore.update(xbaddr, vals)
      aeval(e, new_env, new_store, time)

@@

The function @analyze@ is also changed by adding a @reset@ around the function call to @aeval@.
The afterwards updating on cache becomes sequential.

\begin{lstlisting}
def analyze(e: Expr) = {
  def iter(cache: Cache): Ans = {
    reset {
      val Ans(vss, anscache) = aval(e, Map[String, BAddr](), Store[BAddr, Storable](Map()), List(), cache)
      val initConfig = Config(e, Map[String, BAddr](), Store[BAddr, Storable](Map()), List())
      val new_cache = anscache.outUpdate(initConfig, vss)
      if (new_cache.out == cache.out) { Ans(vss, new_cache) }
      else { iter(Cache(new_cache.out, new_cache.out)) }
    }
  }
  iter(mtCache)
}
\end{lstlisting}

Now we have arrive the end of this series of transformations. Starting from an abstract abstract machine, 
eventually we obtain an abstract interpreter with pushdown control-flows and written in purely direct-style.

@@

This paper is greatly inspired by~\citeauthor{darais2017abstracting}'s work.
The difference on the surface is that ours works for A-Normal Form $\lambda$-calculus,
instead of plain $\lambda$-calculus, although our work can also be easily adopted to handle a plain
$\lambda$-calculus.
But except of simply showing the final form of abstract definitional interpreters,
we reveal that refunctionalization plays an important role for inheriting the
stack from the metalanguage.
The target of the series of transformations in this paper is abstract definitional
interpreters, and we additionally demonstrate abstract interpreters written direct-style with
delimited controls. 
We use the delimited controls @shift@ and @reset@ that are implemented in Scala \cite{rompf2009implementing}.
The correspondence of delimited controls (as well as continuations) and monads
are well-known \cite{Danvy:1990:AC:91556.91622, wadler1992essence, danvy1992representing, moggi1991notions}.

\section{Conclusion}\label{sec:conclusion}

In this functional pearl, we fill the gap between small-step abstract abstract machines
and big-step abstract definitional interpreters by developing a series of syntactical transformations.
Among these transformations, linearization turns worklist to another layer of continuations; 
refunctionalization converts the first-order data types representing continuations to higher-order 
functions; and finally un-CPS with delimited controls transform the abstract interpreter to a sequence of 
expressions.

The broader ...\todo{conc}
